Co-evolution of lexical meaning \& pragmatic use

### 2017-03-23 paper/paper-v3.tex
paper update

@@

\title{Tracing the cultural evolution of meaning at the semantics-pragmatics interface}

@@

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game theoretic model of the competition between types of language
  users, each endowed with her selection of lexical concepts and her own pragmatic dispositions
  to act on them. Our model traces two evolutionary forces and their interaction: (i)
  fitness-based pressure towards communicative efficiency and (ii) potential learning biases
  during the transfer of linguistic knowledge. We illustrate the model based on a case study on
  scalar implicatures. In this case study learning biases that favor simple semantic
  representations can boost the evolution of more sophisticated pragmatic reasoning types and
  so prevent the lexicalization of scalar implicatures.
\end{abstract}

@@

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved and why it would be so pervasive across natural
languages. Models of language evolution abound. There are simulation-based models studying the
evolution of language in populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents. In contrast, we will here look at language users with a richer
cognitive make-up and their dispositions to act.

Towards this end, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using probabilistic models
of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and sparse learning
data. Considering transmission of linguistic knowledge is important because neither semantic
meanings nor pragmatic usage patterns are directly observable. Instead, language learners have
to infer these unobservables from the observable behavior in which they result. We formalize
this process as a form of Bayesian inference. Our approach thereby contains a well-understood
model of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but
combines it with functional selection, here formalized as the most versatile dynamic from
evolutionary game theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another. As natural languages need to be learnable to survive their
faithful transmission across generations. Even small learning biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

Consider the example of a simple coordination game. Agents are of two types: positive and
negative. If agents of different type play with each other, they obtain a payoff of 0. If
negative meets negative, each receives a payoff of 1. If positive meets positive, they get a
payoff of 2. A population state is completely characterized by the proportion $x$ of
negatives. The fitness of negatives is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$. The
average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation, the
replicator dynamic will then update $x$ to $RD(x) = \nicefrac{x^2}{\Phi(x)}$. The replicator
step is plotted in Figure~\ref{fig:Updates_RMS} as the blue line. Rest points, for which
$RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$. The former are attractors as nearby
points converge to them. Points near $x=\nicefrac{2}{3}$ move towards 0 or 1. This is
schematically pictured in the topmost phase portrait in Figure~\ref{fig:Phase_RD}.

Adding mutation changes the dynamics and its rest points. Let's assume that $Q_{ji} = .9$ when
$j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
plotted as the linear green line in Figure~\ref{fig:Updates_RMS}. It has only one stable rest
point at $x = 0.5$. If we first take the replicator dynamic update and then the mutation update
in sequence, we obtain the replicator mutator dynamic $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 -
  .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in Figure~\ref{fig:Updates_RMS}. The rest
points are at $x=.121$, $x=.903$ and $x=.609$. The former two are attractors (see bottom of
Figure~\ref{fig:Phase_RD}).

@@

    \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \begin{tikzpicture}
      \begin{axis}[xlabel=$x$, ylabel = $x'$,
                   grid=major, width=6.85cm, height = 5cm,
                   legend style = {legend pos = north west},
                   ymin=0,ymax=1, xmin=0,xmax=1]

      \addplot[smooth, color=blue] {x^2 / (3*x^2 - 4*x + 2) + 0.01};
      \addplot[smooth, color=green] {.8*x + .1};

@@

    \vspace*{1cm}

@@

    \caption{Phase portraits for RD and RMD: unstable rest points are hollow, attractors are
      solid.}
        \label{fig:Phase_RD}
    \end{subfigure}

>>>>

\newtheoremstyle{Satz}
   {}                      %Space above
   {1em}                   %Space below
   {\normalfont}           %Body font
   {}                      %Indent amount (empty = no indent,
                           %\parindent = para indent)
   {\normalfont}           %Thm head font
   {.}                     %Punctuation after thm head
   {.8em}                  %Space after thm head: " " = normal interword
                           %space; \newline = linebreak
   {\bfseries\thmname{#1}\thmnote{ (#3)}}
                           %Thm head spec (can be left empty, meaning
                           %`normal')

\theoremstyle{Satz}
\newtheorem{example}{Example}

@@

\title{Co-evolution of lexical meaning \& pragmatic use}

@@

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game theoretic model of the competition between types of language
  users, each endowed with a selection of lexical concepts and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) fitness-based pressure towards communicative efficiency and (ii) potential learning
  biases during the transfer of linguistic knowledge. We illustrate the model based on a case
  study on scalar implicatures. In this case study learning biases that favor simple semantic
  representations can boost the evolution of more sophisticated pragmatic reasoning types and
  so prevent the lexicalization of scalar implicatures.
\end{abstract}

@@

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved and why it would be so pervasive across natural
languages. Models of language evolution abound. There are simulation-based models studying the
evolution of language in populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents. In contrast, we will here look at language users with a richer
cognitive make-up.

We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning
types. The objects of replication and selection are pairs of lexical meanings and general types
of pragmatic behavior, which we represent using probabilistic models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and sparse learning
data. Considering transmission of linguistic knowledge is important because neither semantic
meanings nor pragmatic usage patterns are directly observable. Instead, language learners have
to infer these unobservables from the observable behavior in which they result. We formalize
this process as a form of Bayesian inference. Our approach thereby contains a well-understood
model of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but
combines it with functional selection, here formalized as the most versatile dynamic from
evolutionary game theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another. Natural languages need to be learnable to survive their
faithful transmission across generations. Even small learning biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

\begin{example}
  Consider a simple coordination game. Agents are of two types: positive and negative. If
  agents of different type play with each other, they obtain a payoff of 0. If negative meets
  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
  population state is completely characterized by the proportion $x$ of negatives. The fitness
  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{x^2}{\Phi(x)}$. The update
  function $RD(x)$ of the replicator step is plotted in Figure~\ref{fig:Updates_RMS} as the
  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
  Figure~\ref{fig:Phase_RD}.

  Adding mutation changes the dynamics and its rest points. Let's assume that $Q_{ji} = .9$
  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
  plotted as the linear green line in Figure~\ref{fig:Updates_RMS}. It has only one stable rest
  point at $x = 0.5$ (see Figure~\ref{fig:Phase_RD}). If we first take the replicator step and
  then the mutation step in sequence, we obtain the replicator mutator dynamic
  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
  Figure~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
  former two are attractors (see Figure~\ref{fig:Phase_RD}).
\end{example}

@@

  %% notice that the figure is hacked, because somehow the curve does not come out right under
  %% PGFplots; I don't know why!?!?

    \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \begin{tikzpicture}
      \begin{axis}[xlabel=$x$, ylabel = $x'$,
                   grid=major, width=6.85cm, height = 5.5cm,
                   legend style = {legend pos = north west},
                   ymin=0,ymax=1, xmin=0,xmax=1]

      \addplot[smooth, color=blue] {x^2 / (3*x^2 - 4*x + 2) + 0.019 * x};
      \addplot[smooth, color=green] {.8*x + .1};

@@

    \vspace*{0.35cm}

      \node[draw=black, circle, fill = black, minimum size = 0.25cm, thick]
      (mid) at (0.5,0) {};

      \node[]  (label) at (0.5,0.55) {M};

      \draw[->-=0.5,very thick] (0) -- (mid);

      \draw[->-=0.5,very thick] (1) -- (mid);

    \vspace*{0.35cm}

@@

    \caption{Phase portraits for RD, M and RMD: unstable rest points are hollow, attractors are
      solid.}
        \label{fig:Phase_RD}
    \end{subfigure}
